Using reference models in variable selection \vspace{.1in}

### 2022-02-21 paper/ref_approach_blinded.tex
final version for revision

\usepackage{amssymb,amsmath,amsthm}
\usepackage{newtxtext} % times
\usepackage[scaled=.95]{cabin} % sans serif
\usepackage[varqu,varl]{inconsolata} % typewriter
\usepackage{amsmath}
\usepackage[varg]{newtxmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\usepackage{algorithm2e}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={The reference model approach in variable selection},
            %pdfauthor={Federico Pavone, Juho Piironen, Paul-Christian B\"{u}rkner, Aki Vehtari},
            pdfkeywords={keywords},
            pdfborder={0 0 0},
            breaklinks=true,
            colorlinks=true,
            linkcolor=black,
            citecolor=RoyalBlue,
            filecolor=black,
            urlcolor=RoyalBlue,
          }
\urlstyle{same}  % don't use monospace font for urls
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[shorthands=off,main=american]{babel}
\else
  \usepackage{polyglossia}
  \setmainlanguage[variant=american]{english}
\fi
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{placeins}
\usepackage{graphicx,grffile,subcaption}
\textwidth 6.0 true in
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{3}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

@@

%    \author{
%      Federico Pavone\footnote{Department of Computer Science, Aalto University, Finland.}\, \footnote{Department of Decision Sciences, Bocconi University, Italy.}%$\,$\footnote{Correspondence concerning this article should be addressed to Federico Pavone, Department of Decision Sciences, Bocconi University, Via R\"{o}ntgen 1, 20136 Milano, Italy. E-mail: federicopavone01@gmail.com.}
%      \,,\,
%  Juho Piironen$^*$,\,
%  Paul-Christian B\"{u}rkner$^*$
%  and Aki Vehtari$^*$
%  }

@@

\begin{document}
\maketitle
\begin{abstract}
  Variable selection, or more generally, model reduction is an important aspect of the statistical workflow aiming to provide insights from data. In this paper, we discuss and demonstrate the benefits of using a reference model in variable selection. A reference model acts as a noise-filter on the target variable by modeling its data generating mechanism. As a result, using the reference model predictions in the model selection procedure reduces the variability and improves stability leading to improved model selection performance. Assuming that a Bayesian reference model describes the true distribution of future data well, the theoretically preferred usage of the reference model is to project its predictive distribution to a reduced model leading to projection predictive variable selection approach. Alternatively, reference models may also be used in an ad-hoc manner in combination with common variable selection methods.  In several numerical experiments, we investigate the performance of the projective prediction approach as well as alternative variable selection methods with and without reference models. Our results indicate that the use of reference models generally translates into better and more stable variable selection. Additionally, we demonstrate that the projection predictive approach shows superior performance as compared to alternative variable selection methods independently of whether or not they use reference models.
\end{abstract}

@@

In data-based approaches, such as Lasso selection
\citep{tibshirani1996regression} or stepwise backward/forward
regression \citep{venables2013modern,harrell2015regression}, the
observed empirical data distribution is utilised as a proxy of future
data usually in combination with cross-validation or information
criteria to provide estimates for out-of-sample predictive
performance.  In contrast, reference model based methods approximate
the future data generation mechanism using the predictive distribution
of a reference model, which can be, for example, a full-encompassing
model including all variables.  The main assumption under any
reference model approach is that we operate in an
$\mathcal{M}$-\textit{complete} framework
\citep{book:bernardo_smith,vehtari2012survey}, that is, we have
constructed a model which reflects our beliefs about
the future data in the best possible way and which has passed model checking and criticism
\citep[see, e.g.][]{gelman2013bayesian,gabry2019visualization}.  The reference model approach
has been used in Bayesian statistics at least since the seminal
work of \citet{paper:reference_lindley}. For more historical
references, see \citet{vehtari2012survey} and
\citet{paper:model_selection}, and for most recent methodological
developments see \citet{paper:projpred}.

Reference models have been also used in non-Bayesian contexts, in which
\cite{harrell2015regression} describes them as full models that can be
thought of as a ``gold standard'' (for a given application).  For
example, \cite{faraggi2001understanding} deal with the necessity of
identifying interpretable risk groups in the context of survival data
using neural networks, which typically perform very well in terms of
prediction, but whose variables are difficult to be understood in
terms of relevance.  \cite{paul2008preconditioning}, using the term
preconditioning, explore approximating models fitting Lasso or
stepwise regression against consistent estimates $\hat{y}$ of a
reference model instead of the observed responses $y$.
Whatever the terminology or applied statistical framework, reference
models offer a powerful approach to improving variable selection as we
will demonstrate in the present paper.

The main contributions of our work are:

\begin{itemize}
\item We show a simple and intuitive way to combine any variable selection method with the reference model approach.
\item We perform extensive numerical experiments to compare variable selection methods with or without using a reference model, both for complete and minimal subset feature selection.
\item We provide evidence supporting the projection predictive approach as a principled way to use reference models.
\end{itemize}

The paper is structured as follows. In Section
\ref{ref-intro}, we review the concept of the reference
model, its benefits with examples and how it can be used as a filter
on data in a simple way. In Section \ref{comparison-minimal-subset} and 
Section \ref{comparison-complete-subset}, we show the
benefits of a reference model approaches for minimal and complete 
variable selection, respectively, before we end with a conclusion in Section \ref{conclusion}.

@@

To motivate the further discussion and experiments, we start by a simple
variable selection example using body fat data by \citet{johnson1996fitting}.
We compare the projective prediction approach (\emph{projpred},
\citet{paper:projpred}) which uses a reference model, and classic
stepwise backward regression (\emph{steplm}).
The experiments are implemented in \texttt{R} \citep{Rcore2018}.

The target variable of interest is the amount of body fat, which is
obtained by a complex and expensive procedure consisting in immersing a 
person in a water tank and carrying out different measurements and computations. 
Additionally, we have information about 13
variables which are anthropometric measurements (e.g., height, weight
and circumference of different body parts). The
variables are highly correlated which causes additional
challenge in the variable selection. In total, we have 251
observations. The goal is to find the model which is able to predict
the amount of body fat well while requiring the least amount of
measurements for a new person.

\cite{paper:bodyfat} report results using steplm with a significance
level of 0.157 with AIC selection \citep{akaike1974new}, fixing
abdomen and height to be always included in the model. For better
comparison we do not fix any of the variables.  The steplm approach is
carried out combining the \texttt{step} and \texttt{lm} functions in \texttt{R}.

For the selection via projpred, the Bayesian reference model includes
all the variables using a regularised horseshoe prior~\citep{paper:rhs} on the
variable coefficients. Submodels are explored using
forward search (the results are not sensitive to whether forward or
backward search is used), and the predictive utility is the expected
log-predictive density (elpd) estimated using approximate leave-one-out cross-validation via Pareto-smoothed importance-sampling
\citep[PSIS-LOO-CV; ][]{paper:psis_loo}.  We select the smallest  submodel with an elpd score similar to the reference model
when taking into account the uncertainty in estimating the predictive
model performance. See Appendix A for brief review of projection
predictive approach, and papers by \citet{paper:model_selection} and
\citet{paper:projpred} for more details. The complete projpred approach is implemented in the \texttt{projpred} \texttt{R} package \citep{Rprojpred}.

@@

Both projpred and steplm compare a large number of models using either
forward or backward search, which can lead to selection induced
overfitting, but even with 100 variables, projpred is able to select a
submodel with similar performance as the full model. In this example, the two
 compared methods also differ in other aspects than the usage of a reference 
 model, such as that projpred
uses Bayesian inference and steplm uses maximum likelihood estimation. 
However, as we show in the present paper, one of the primary difference 
with respect to quality of the variable selection is indeed whether or 
not a reference model is applied.

@@

A properly designed reference model is able to filter parts of the
noise present in the data, and hence to provide an improved and more
stable selection process. This holds even if the reference model does
not perfectly resemble the true data generating process. Moreover, our
analyses indicate that the substantial reduction of variance
attributable to noise is usually more important than small potential
bias due to model misspecification.  We argue that, regardless of how
the reference model is set up and used in the inference procedure, it
can be always seen as acting as a filter on the observed
data. Furthermore, regardless of what specific model selection method
is used, a reference model can be used instead of raw data during the
selection process to improve the stability and selection
performance. Our results indicate that the core reason why the
reference model based methods perform well is the reference model
itself, rather than the specific way of using it.  In general, the
less data we have and the more complex the estimated models are, the
higher is the benefit of using a reference models as the basis for
variable selection.

If one of the models to be compared is the full model, which can be
used as a reference model, there is no additional cost of using a
reference model as it was estimated as part of the analysis anyway.
Sometimes including all the available variables in an
elaborate model can be computationally demanding.  In such a
case, even simpler screening or dimensionality reduction techniques,
as for example the supervised principal components
\citep{bair2006prediction,piironen2018} can produce useful reference
models.

@@

When using a reference model, we first obtain predictions for $f$
using all the variables $\{X_{j}\}_{j=1}^{p}$ taking into account
that we have only observed the noisy representation $Y$ of $f$. If our
model is good, we are able to describe $f$ better than $Y$ itself can,
which improves the accuracy in the estimation of the relevance of the
variables.  Figure \ref{fig:correlation} illustrates this process in
the form of a scatter plot of (absolute) correlations of the
variables with $Y$ against the corresponding correlations with the
predictions of a reference model (in this case the posterior
predictive means of model \eqref{eq:ref_mod}; see Section
\ref{simulations}). Looking at the marginal distributions, we see
that using a reference model to filter out noise in the data, the two
groups of variables (relevant and non-relevant) can be distinguished
much better than when the correlation is computed using the observed
noisy data directly.
\begin{figure}[tp]
  \centering
  \vspace{-5mm}
  \includegraphics[width=0.35\textwidth]{graphics/correlation.pdf}
  \vspace{-0.5\baselineskip}
  \caption{Sample correlation plot of each variable (relevant in red, 
  non-relevant in blue) with the target variable $y$ and the latent variable 
  $f$ respectively on the x and y axis. Simulated data are generated according to 
  \eqref{eq:simulated_data} with parameters $n=70$, $\rho=0.3$,
  $k=100$, $p=1000$.}
  \label{fig:correlation}
\end{figure}

@@

In the body fat example above, our two simultaneous goals were to 
obtain good predictive performance and to select a smaller number of variables. 
When the goal is to select a minimal subset of variables,
which have similar predictive performance as the full model, we call it
\emph{minimal subset variable selection}. This minimal subset might
exclude variables which have some predictive information about the
target but, given the minimal subset, these variables are not able to
provide such additional information that would improve predictive
performance in a substantial manner.
The usual reason for this is that the relevant variables
which are not in the minimal subset are highly correlated with variables already in
the minimal subset. We will return to a problem of finding all
the variables with some predictive power in Section~\ref{complete-selection}.

\subsection{Simulation study 1}

Using the data generating mechanism \eqref{eq:simulated_data}, we
simulate data sets of different sizes with a relatively large number
of variables $p=70$, with $k=20$ of them being predictive.  We compare
the minimal subset variable selection performance of the projection
predictive approach (which uses a reference model and it is referred to as projpred), 
a Bayesian stepwise
forward selection with and without a reference model, and maximum
likelihood stepwise forward selection with and without a reference
model (steplm). The following is a summary of the implementation of the compared methods:
\begin{itemize}
\item \textbf{projpred}: the projective prediction approach is used. 
  The reference model is a Bayesian linear regression
  model using the first five supervised principal components \citep{piironen2018} as predictors
  and the full posterior predictive distribution as the basis of the projection.   
  The search heuristic  is forward search and the predictive performance 
  is estimated via 10-fold cross-validation. The selection continues
  until the predictive performance is close to the predictive
  performance of the reference model. See more details in Appendix A.

@@

In Figure \ref{fig:rmse_vs_fdr}, the predictive performance is shown
for different values of $n$ and $\rho$ in terms of RMSE and the false discovery rate (FDR, the ratio of the number of non-relevant selected
variables over the number of selected variables) of the selected submodel,
averaged after 100 data simulations. We see that projpred has superior FDR and the
smallest RMSE. Using a reference model improves steplm significantly,
but the minimal subset variable selection stays unreliable. Stepwise
Bayesian linear regression is better than steplm when comparing both methods with
and without a reference model, respectively. However, the minimal subset variable
selection of the Bayesian linear regression is still less reliable than projpred.

When the variable selection is repeated with different simulated data
sets, there is some variability in the selected variables. We
measure the stability of variable selection by computing the entropy
of the observed distribution of the included variables over different models. The smallest entropy
would be obtained if the approach always selected the same set of
variables, and the largest entropy would be observed if the approach
would always selected different sets of variables. Therefore, lower entropy 
corresponds to a more stable selection. Highly correlated predictive variables may happen to be
selected alternately, thus making stability estimation of the
selection a non-trivial task. Entropy can not distinguish the interchangeability 
due to correlation from instability. Thus, such a measure should be 
considered as a relative, and not as an absolute, measure of stability.
Figure \ref{fig:entropy} shows the entropy scores for the different compared
methods. The use of a reference model
improves the stability of steplm in variable selection slightly, while it makes
little difference for the Bayesian linear regression. The
projpred approach turns out to be far more stable than all
other methods. This is likely due to projpred being based on better
decision theoretical formulation which 1) takes into account the full
predictive distribution and not just point estimate and 2) projects
the reference model posterior to the submodel instead of using a
simple refit of submodels.

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/rmse_vs_fdr_parallel.pdf}
  \vspace{-0.3\baselineskip}
  \caption{Simulation study 1: Root mean square error (RMSE) against false discovery rate in the minimal subset variable selection with one standard deviation error bars. The projpred approach has the smallest RMSE and false discovery ratio.}
  \label{fig:rmse_vs_fdr}
\end{figure}

@@

subset. Than, the main steps are the following:

>>>>

\usepackage{amssymb,amsmath,amsthm}
\usepackage{newtxtext} % times
\usepackage[scaled=.95]{cabin} % sans serif
\usepackage[varqu,varl]{inconsolata} % typewriter
\usepackage{amsmath}
\usepackage[varg]{newtxmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\usepackage{algorithm2e}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={The reference model approach in variable selection},
            pdfauthor={Federico Pavone, Juho Piironen, Paul-Christian B\"{u}rkner, Aki Vehtari},
            pdfkeywords={keywords},
            pdfborder={0 0 0},
            breaklinks=true,
            colorlinks=true,
            linkcolor=black,
            citecolor=RoyalBlue,
            filecolor=black,
            urlcolor=RoyalBlue,
          }
\urlstyle{same}  % don't use monospace font for urls
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[shorthands=off,main=american]{babel}
\else
  \usepackage{polyglossia}
  \setmainlanguage[variant=american]{english}
\fi
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{placeins}
\usepackage{graphicx,grffile,subcaption}
\textwidth 6.0 true in
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{3}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

% \newcommand{\fede}[1]{\textcolor{red}{#1}}

@@

    %\author{
   %   Federico Pavone\footnote{Department of Computer Science, Aalto University, Finland.}\, \footnote{Department of Decision Sciences, Bocconi University, Italy.}%$\,$\footnote{Correspondence concerning this article should be addressed to Federico Pavone, Department of Decision Sciences, Bocconi University, Via R\"{o}ntgen 1, 20136 Milano, Italy. E-mail: federicopavone01@gmail.com.}
 %     \,,\,
 % Juho Piironen$^*$,\,
 % Paul-Christian B\"{u}rkner$^*$
 % and Aki Vehtari$^*$
 % }

@@

\begin{document}
\maketitle
\begin{abstract}
  Variable selection, or more generally, model reduction is an important aspect of the statistical workflow aiming to provide insights from data. In this paper, we discuss and demonstrate the benefits of using a reference model in variable selection. A reference model acts as a noise-filter on the target variable by modeling its data generating mechanism. As a result, using the reference model predictions in the model selection procedure reduces the variability and improves stability, leading to improved model selection performance. Assuming that a Bayesian reference model describes the true distribution of future data well, the theoretically preferred usage of the reference model is to project its predictive distribution to a reduced model, leading to projection predictive variable selection approach. We analyse how much the great performance of the projection predictive variable is due to the use of reference model and show that other variable selection methods can also be greatly improved by using the reference model as target instead of the original data. In several numerical experiments, we investigate the performance of the projective prediction approach as well as alternative variable selection methods with and without reference models. Our results indicate that the use of reference models generally translates into better and more stable variable selection.
  %Additionally, we demonstrate that the projection predictive approach shows superior performance as compared to alternative variable selection methods, independently of whether or not they use reference models.
\end{abstract}

@@

In data-based approaches, such as Lasso selection
\citep{tibshirani1996regression} or stepwise backward/forward
regression \citep{venables2013modern,harrell2015regression}, the
observed empirical data distribution is utilised as a proxy of future
data, usually in combination with cross-validation or information
criteria to provide estimates of out-of-sample predictive
performance.  In contrast, reference model based methods approximate
the future data generation mechanism using the predictive distribution
of a reference model, which can be, for example, a full-encompassing
model including all variables.

We assume all models are wrong, but we assume we have constructed a
model which reflects our beliefs about the future data in the best
possible way and which has passed model checking and criticism
\citep{gelman2020bayesian,gabry2019visualization}.  Using the usual
best practices for constructing the reference model is important, as
using a bad reference model can only lead to selecting a similarly
bad smaller model. If the reference model is considered to have
useful predictions, then the smaller models selected will also have
similar useful predictions.
%
The reference model approach has been used in Bayesian statistics in
some form at least since the seminal work of
\citet{paper:reference_lindley}. For more historical references, see
\citet{vehtari2012survey} and \citet{paper:model_selection}, and for
most recent methodological developments see \citet{paper:projpred}.
%
Examples of useful reference models can be found for example for
small-$n$-large-$p$ regression and logistic regression by
\citet{piironen2017comparison} (with spike-and-slab prior),
\citet{Piironen:2015:Stan_projection} (with horseshoe prior), and
\citet{paper:projpred} (with iterative supervised principal
components), for generalized linear and additive multilevel models by
\citet{Catalina:2020:GAMM_projection}, for regression models with
non-exponential family observation models by
\citet{Catalina+etal:2021:latent_projection}, and for generic
multivariate non-linear regression with higher order interactions by
\citet{Piironen:2016:GPprojection}.

Reference models have been also used in non-Bayesian context.
\cite{harrell2015regression} describes them as full models that can be
thought of as a ``gold standard'' (for a given application).
\cite{faraggi2001understanding} deal with the necessity of
identifying interpretable risk groups in the context of survival data
using neural networks, which typically perform very well in terms of
prediction, but whose variables are difficult to be understood in
terms of relevance.  \cite{paul2008preconditioning}, using the term
preconditioning, explore approximating models fitting Lasso or
stepwise regression against consistent estimates $\hat{y}$ of a
reference model instead of the observed responses $y$.

All these methods can be framed into the family of reference model approaches. 
The common denominator is the use of the reference model predictive information 
instead of simply observed data to guide the selection.
Whatever the terminology or applied statistical framework, reference
models offer a powerful approach to improving variable selection, as we
will demonstrate in the present paper.

The goal of the present paper is to study the impact of reference model approaches by disentangling the benefit of using reference models per se from the benefit of specific variable selection algorithms. In particular, we:
\begin{itemize}
	\item propose a simple and intuitive approach to combine any variable selection method with the reference model approach. This allows us to investigate the benefit of using reference models independent of the specific variable selection method;
	\item perform extensive numerical experiments to compare variable selection methods with or without using a reference model, both for complete and minimal subset variable selection and assessing the quality of the selection;
	\item provide evidence supporting, in particular, the projection predictive approach as a principled way to use reference models in minimal subset variable selection.
\end{itemize}

The paper is structured as follows. In Section
\ref{ref-intro}, we review the concept of the reference
model, its benefits with examples and how it can be used as a filter
on data in a simple way. In Section \ref{comparison-minimal-subset} and 
Section \ref{comparison-complete-subset}, we show the
benefits of a reference model approaches for minimal and complete 
variable selection, respectively, before we end with a conclusion in Section \ref{conclusion}.
The code to run all
the experiments is available on GitHub\footnote{\url{https://github.com/fpavone/ref-approach-paper}}.

@@

To motivate the further discussion and experiments, we start by a simple
variable selection example using body fat data by \citet{johnson1996fitting}. The same data was used illustrate the variable selection with classic stepwise backward regression by \citet{paper:bodyfat}.
We compare the projective prediction approach (\emph{projpred},
\citet{paper:projpred}) which uses a reference model, and classic
stepwise backward regression (\emph{steplm}).
The experiments are implemented in \texttt{R} \citep{Rcore2018}.

The target variable of interest is the amount of body fat, which is
obtained by a complex and expensive procedure consisting in immersing a 
person in a water tank and carrying out different measurements and computations. 
Additionally, we have information about 13
variables which are anthropometric measurements (e.g., height, weight
and circumference of different body parts). The
variables are highly correlated, which causes additional
challenge in the variable selection. In total, we have 251
observations. The goal is to find the model which is able to predict
the amount of body fat well while requiring the least amount of
measurements for a new person.

\cite{paper:bodyfat} report results using steplm with a significance
level of 0.157 with AIC selection \citep{akaike1974new}, fixing
abdomen and height to be always included in the model. For better
comparison, we do not fix any of the variables.  The steplm approach is
carried out combining the \texttt{step} and \texttt{lm} functions in \texttt{R}.

For the selection via projpred, the Bayesian reference model includes
all the variables using a regularised horseshoe prior~\citep{paper:rhs} on the
variable coefficients. Submodels are explored using
forward search (the results are not sensitive to whether forward or
backward search is used), and the predictive utility is the expected
log-predictive density (elpd) estimated using approximate leave-one-out cross-validation via Pareto-smoothed importance-sampling
\citep[PSIS-LOO-CV; ][]{paper:psis_loo}.  We select the smallest  submodel with an elpd score similar to the reference model
when taking into account the uncertainty in estimating the predictive
model performance. See Appendix A for a brief review of projection
predictive approach, and papers by \citet{paper:model_selection} and
\citet{paper:projpred} for more details. The complete projpred approach is implemented in the \texttt{projpred} \texttt{R} package \citep{Rprojpred}.

@@

Both projpred and steplm compare a large number of models using either
forward or backward search, which can lead to selection induced
overfitting, but even with 100 variables, projpred is able to select a
submodel with similar performance as the full model. In this example, the two
compared methods also differ in other aspects than the usage of a reference 
model, such as that projpred
uses Bayesian inference and steplm uses maximum likelihood estimation.
%
To separate the effect of using a reference model, we show that
performance of other variable selection methods, including steplm, can
also be improved by using reference models.
% However, as we show in the present paper, one of the primary difference 
% with respect to quality of the variable selection is indeed whether or 
% not a reference model is applied, and the performance of steplm can
% also be improved by using a reference model.

@@

A properly designed reference model is able to filter parts of the
noise present in the data, and hence to provide an improved and more
stable selection process. This holds even if the reference model does
not perfectly resemble the true data generating process. 
Clearly, the reference model approach requires a sensible model. 
The construction of such a model should follow proper modelling workflow \citep[see, e.g.][]{gelman2020bayesian}. Better predictive models imply better selection results. 
The goodness of a reference model comes from its predictive ability which should be 
assessed via proper validation methods. Our analyses indicate that the substantial 
reduction of variance attributable to noise is usually more important than small potential
bias due to model misspecification.
When the goal is ranking the variables for importance to
perform variable selection, it is known that large variance is
generally more harmful than small bias \citep{piironen2017comparison}.

In general, there is no restriction on the type of model the reference models should be, and a sensible reference model does not even need to be Bayesian necessarily.
However, Bayesian methods can help in some of those situations
where MLE procedures struggle. 
It is known that as the number of parameters in the model increases, the MLE
estimator is dominated by shrinkage estimators 
\citep{stein1956inadmissibility, stein1961estimation, parmigiani2009decision, efron2011tweedie}.
The use of a prior in Bayesian inference automatically incorporates 
some kind of shrinkage in the Bayesian estimator under a given loss function \citep[see, e.g.][]{rockova2012hierarchical}. 
For the sake of our study, we rely on simple data structure which can be well described 
by linear regression models. However, more complex data can arise in practice, for example, including hierarchical structures. In these cases, MLE inference tends to become cumbersome, whereas the Bayesian framework provides a natural way to convey uncertainties and make inference through the joint posterior distribution of all parameters.

We argue that, regardless of how
the reference model is set up and used in the inference procedure, it
can be always seen as acting as a filter on the observed
data. Furthermore, regardless of what specific model selection method
is used, a reference model can be used instead of raw data during the
selection process to improve the stability and selection
performance. Our results indicate that the core reason why the
reference model based methods perform well is the reference model
itself, rather than the specific way of using it.  In general, the
less data we have and the more complex the estimated models are, the
higher is the benefit of using a reference models as the basis for
variable selection.

If one of the models to be compared is the full model, which can be
used as a reference model, there is no additional cost of using a
reference model as it was estimated as part of the analysis anyway.
Sometimes, including all the available variables in an
elaborate model can be computationally demanding.  In such a
case, even simpler screening or dimensionality reduction techniques,
as for example the supervised principal components
\citep{bair2006prediction,piironen2018}, can produce useful reference
models \citep{paper:projpred}. As an example, Table \ref{tab:computational_time} reports
the computational time for the body fat data example of Section \ref{bodyfat-1}. 
For reference model approaches, i.e. projpred and ref+step.lm, the main
time consuming operation is to fit the reference model. The burden of it depends clearly
on the specific application and type of model used.

\begin{table}[tp]
	\footnotesize
	\centering
	\begin{tabular}{l|cc|ccc}
		Data & Reference model & Search path & projpred & step.lm & ref + step.lm \\
		\hline
		body fat & 38 & 15 & 53 &  &  \\
		  &  & 0.08 &  & 0.08 &  \\
		  & 38 & 0.08 &  &  & 38 \\
		\hline 
		+ noisy variables & 170 & 78 & 247 &  &  \\
		  & & 12 & & 12 & \\
		  & 170 & 12 & & & 182

	\end{tabular}
	\caption{Computational time (in seconds) of body fat example of Section \ref{bodyfat-1}.
		The reported times for each each method on the three last columns are sums of the time to fit the reference model (second column) and
		to explore the search path (third column).}
	\label{tab:computational_time}
\end{table}

@@

When using a reference model, we first obtain predictions for $f$
using all the variables $\{X_{j}\}_{j=1}^{p}$ taking into account
that we have only observed the noisy representation $Y$ of $f$. If our
model is good, we are able to describe $f$ better than $Y$ itself can,
which improves the accuracy in the estimation of the relevance of the
variables.  Figure \ref{fig:correlation} illustrates this process in
the form of a scatter plot of (absolute) correlations of the
variables with $Y$ against the corresponding correlations with the
predictions of a reference model (in this case, the posterior
predictive means of model \eqref{eq:ref_mod}; see Section
\ref{simulations}). Looking at the marginal distributions, we see
that using a reference model to filter out noise in the data, the two
groups of variables (relevant and non-relevant) can be distinguished
much better than when the correlation is computed using the observed
noisy data directly.
\begin{figure}[tp]
  \centering
  \vspace{-5mm}
  \includegraphics[width=0.35\textwidth]{graphics/correlation.pdf}
  \vspace{-0.5\baselineskip}
  \caption{Sample correlation plot of each variable (relevant in red, 
  non-relevant in blue) with the target variable $y$ and the latent variable 
  $f$ respectively on the x- and y-axes. Simulated data are generated according to 
  \eqref{eq:simulated_data} with parameters $n=70$, $\rho=0.3$,
  $k=100$, $p=1000$.}
  \label{fig:correlation}
\end{figure}

@@

In the body fat example above, our two simultaneous goals were to 
obtain good predictive performance and to select a smaller number of variables. 
When the goal is to select a minimal subset of variables,
which have similar predictive performance as the full model, we call it
\emph{minimal subset variable selection}. This minimal subset might
exclude variables which have some predictive information about the
target but, given the minimal subset, these variables are not able to
provide such additional information that would improve predictive
performance in a substantial manner.
The usual reason for this is that the relevant variables
which are not in the minimal subset are highly correlated with variables already in
the minimal subset. Such nature of the minimal subset makes the solution
not unique, except in the particular case of completely orthogonal predictors. 
The non-uniqueness of the solution is not a problem per se, as different samples of the data
are expected to give possible different minimal subsets but with same predictive power. 
However, it makes it difficult to define a proper
concept of stability of the selection.
We will return to a problem of finding all
the variables with some predictive power in Section~\ref{complete-selection}.
%
% The code to run all
% the experiments is available on GitHub\footnote{\url{https://github.com/fpavone/ref-approach-paper}}.

\subsection{Simulation study 1}\label{minimal-simulated}

Using the data generating mechanism \eqref{eq:simulated_data}, we
simulate data sets of different sizes with a relatively large number
of variables $p=70$, with $k=20$ of them being predictive.  We compare
the minimal subset variable selection performance of the projection
predictive approach (which uses a reference model and it is referred to as projpred), 
a Bayesian stepwise
backward selection with and without a reference model, and maximum
likelihood stepwise backward selection with and without a reference
model (steplm). The following is a summary of the implementation of the compared methods:
\begin{itemize}
\item \textbf{projpred}: the projective prediction approach is used. 
  The reference model is a Bayesian linear regression
  model using the first five supervised principal components \citep{piironen2018} as predictors
  and the full posterior predictive distribution as the basis of the projection.   
  The search heuristic  is forward search and the predictive performance 
  is estimated via 10-fold cross-validation. The selection continues
  until the predictive performance is close to the predictive
  performance of the reference model. See more details in Appendix A.

@@

In Figure \ref{fig:rmse_vs_fdr}, the predictive performance is shown
for different values of $n$ and $\rho$ in terms of RMSE and the false discovery rate (FDR, the ratio of the number of non-relevant selected
variables over the number of selected variables) of the selected submodel,
averaged after 100 data simulations. Use of a reference models greatly reduces the overfitting in the selection process, and thus produces much lower test set RMSE of the selected model for all the methods. projpred has the lowest FDR and RMSE. Using a reference model improves also FDR of steplm stepwise Bayesian liena rgegression significantly.

When the variable selection is repeated with different simulated data
sets, there is some variability in the selected variables. We
measure the stability of variable selection by computing the entropy
of the observed distribution of the included variables over different models. The smallest entropy
would be obtained if the approach always selected the same set of
variables, and the largest entropy would be observed if the approach
would always select different sets of variables. Therefore, lower entropy 
corresponds to a more stable selection. Highly correlated predictive variables may happen to be
selected alternately, thus making stability estimation of the
selection a non-trivial task. Entropy can not distinguish the interchangeability 
due to correlation from instability. Thus, such a measure should be 
considered as a relative, and not as an absolute, measure of stability.
Figure \ref{fig:entropy} shows the entropy scores for the different compared
methods. The use of a reference model
improves the stability of steplm in variable selection slightly, while it makes
little difference for the Bayesian linear regression. The
projpred approach turns out to be far more stable than all
other methods. This is likely due to projpred being based on a better
decision theoretical formulation which 1) takes into account the full
predictive distribution and not just a point estimate and 2) projects
the reference model posterior to the submodel instead of using a
simple refit of submodels.

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/rmse_vs_fdr_parallel.pdf}
  \vspace{-0.3\baselineskip}
  \caption{Simulation study 1: Root mean square error (RMSE) against false discovery rate in the minimal subset variable selection with one standard deviation error bars. Use of a reference model reduces RMSE of the selected model for all the methods. The projpred approach has the smallest RMSE and false discovery ratio.}
  \label{fig:rmse_vs_fdr}
\end{figure}

@@

subset. Then, the main steps are the following:
